Differential invariant signatures for planar Lie group transformations of images

### 2020-11-17 paper/newpaper.tex
missed one
Similarity and E(2) done

The function is depicted as a greyscale image in Figure~\ref{fig:function}.
On the left the image itself is shown, while on the right a version with a
regular grid superimposed is given. This is useful for visualising both the
transformation and how the image is mapped into the three-dimensional
signature surface.

@@

We could explicitly write the calculation in Eq.~\ref{eq:prolongation} for
this case, but instead provide an intuitive explanation. Consider a point
in the original image, and the corresponding point in the transformed one.
Clearly, the intensity value of the points match, and so do their first
spatial derivatives. These three quantities can be used as coordinates in a
new space, where the two image points will correspond. Computing these
coordinates for all image points will define a surface in the new
three-dimensional space, and the two images will produce the same surface,
which we call the signature.

Figure~\ref{fig:T2} shows (a) the original image, (b) the translated image
under the transformation $(x,y) \mapsto (x + 0.1, y - 0.2)$, (c) the
signature of one of them, (d) the superimposed pair of signatures. The
yellow lines in (c) correspond to where the yellow grid from (a) has been
mapped to on the signature surface. In (d), notice that as some of the
image has moved out of frame, there are parts of the signature near its
boundary that do not overlap.

@@

The group that has been of particular interest for image analysis is the projective group, because of its ubiquity in camera models. Summarise these...
\cite{Suk2000,Suk2004,Hann2002a,Kogan2014}
\cite{Gros1992,Hubert2007}

@@

In this setting, a Euclidean transformation acting on $\mathbb{R}^2$ is given
by an expression of the form:
\begin{equation*}
  \bar{x}_i = a_{ij}x_j + b_j,
\end{equation*}
using the Einstein summation convention. The orthogonality condition is that $a_{ik}a_{jk} = a_{ki}a_{kj} = \delta_{ij}$.

@@

The main observation is that in this setting, the $n^\text{th}$ order partial derivative operator $\partial^n / \partial_{x_{i_1}}\cdots \partial_{x_{i_n}}$ formally transforms as an $n$-tensor~\cite{Florack1993}. Because the product of a $p$-tensor and a $q$-tensor gives a $(p+q)$-tensor and contraction of a $p$-tensor yields a $(p-2)$-tensor, a complete contraction of a tensor of even rank formed as a product of terms of the form $\partial^p f / \partial_{x_{i_1}} \cdots \partial_{x_{i_p}}$ will give a Euclidean invariant.

For example, up to second derivatives, we have the following Euclidean invariants, expressed as tensor contractions:
\begin{align*}
  f &= f \\
  f_if_i &= f_x^2 + f_y^2 \\
  f_{ii} &= f_{xx} + f_{yy} \\
  f_{ij} f_{ij} &= f_{xx}^2 + 2f_{xy}^2 + f_{yy}^2 \\
  f_i f_j f_{ij} &= f_x^2 f_{xx} + 2f_x f_y f_{xy} + f_y^2 f_{yy}
\label{eq:se2}
\end{align*}

While we can construct any number of invariants this way, there is no guarantee of functional independence between them. We will address this issue in Section~\ref{sec:movingframes} when we construct invariants of $E(2)$ and other groups using the method of moving frames.

SE(2): rotation, Sim(2): zooming.

\begin{figure}
\centering
\includegraphics[width=.45\textwidth]{Figs/f_transformed_SE2.png}
\includegraphics[width=.45\textwidth]{Figs/SE2_signature.png}
\caption{A sample $SE(2)$ Transform and 3D signature.}
\label{fig:SE2}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=.45\textwidth]{Figs/f_transformed_E2.png}
\includegraphics[width=.45\textwidth]{Figs/E2_signature.png}
\caption{A sample $E(2)$ Transform and 3D signature.}
\label{fig:E2}
\end{figure}

@@

\subsection{The Special Euclidean Group SE(2)}

@@

\subsection{The Euclidean Group E(2)}
An element of the Euclidean group $E(2)$ maps an element $(x, y)$ of
$\mathbb{R}^2$ to $(\bar{x}, \bar{y})$ as:
\begin{equation*}
  \begin{aligned}
    \bar{x} &=  x\varepsilon\cos\theta  - y\varepsilon\sin\theta + t_x \\
    \bar{y} &= x\sin\theta  + y\cos\theta + t_y.
  \end{aligned}
\end{equation*}
where $\varepsilon \in \{-1, 1\}$ and the remaining parameters are the same
as for $SE(2)$. If $\varepsilon = 1$ it is a rigid transformation, and if
$\varepsilon = -1$ it contains a reflection.

>>>>

The function is depicted as a greyscale image in Figure~\ref{fig:function}.
On the left the image itself is shown, while on the right a version with a
regular grid superimposed is given. We will use this function throughout,
using the grid to help visualise both the transformation and how the image
is mapped into the three-dimensional signature surface.

@@

We could explicitly write the calculation in Eq.~\ref{eq:prolongation} for
this case, but instead provide an intuitive explanation. Consider a point
in the original image, and the corresponding point in the transformed one.
Clearly, the intensity value of the points match, and so do their first
spatial derivatives. These three quantities can be used as coordinates in a
new space, where the two image points will correspond. Computing these
coordinates for all image points will define a surface in the new
three-dimensional space, and the two images will produce the same surface,
which we call the signature, $\mathcal{I}(f)(x, y) = \left(I_1, I_2,
I_3\right)$ where $I_1 = f, I_2 = f_x, I_3 = f_y$.

Figure~\ref{fig:T2} shows (a) the original image, (b) the translated image
under the transformation $(x,y) \mapsto (x + 0.1, y - 0.2)$, (c) the
signature of one of them, (d) the superimposed pair of signatures. The
yellow lines in (c) correspond to where the yellow grid from (a) has been
mapped to on the signature surface. The parts of the surface where the mesh
gets close together are where the function and its derivatives are tending
towards zero. This occurs around the boundary of our test image and so all
of these points are close together in the signature space. In (d), notice
that as some of the image has moved out of frame, there are parts of the
signature surfaces of the two images that do not overlap.

@@

The group that has been of particular interest for image analysis is the projective group, because of its ubiquity in camera models. Summarise these...
\cite{Suk2000,Suk2004,Hann2002a,Kogan2014,Li19}
\cite{Gros1992,Hubert2007}

@@

In this setting, a Euclidean transformation acting on $\mathbb{R}^2$ is given
by an expression of the form:
\begin{equation*}
  \bar{x}_i = a_{ij}x_j + b_j,
\end{equation*}
using the Einstein summation convention. The orthogonality condition is
that $a_{ik}a_{jk} = a_{ki}a_{kj} = \delta_{ij}$. A Euclidean
transformation $\varphi \colon \mathbb{R}^2 \to \mathbb{R}^2$ can also be
written in matrix form as $\bar{\mathbf{x}} = \varphi(\mathbf{x}) =
U\mathbf{x} + \mathbf{t}$ where the matrix $U$ satisfies $U^TU = I$ and
$\mathbf{t} \in \mathbb{R}^2$.

@@

The main observation is that in this setting, the $n^\text{th}$ order
partial derivative operator $\partial^n / \partial_{x_{i_1}}\cdots
\partial_{x_{i_n}}$ formally transforms as an $n$-tensor, \citep{Florack1993}. Because the product of a $p$-tensor and a $q$-tensor gives a $(p+q)$-tensor and contraction of a $p$-tensor yields a $(p-2)$-tensor, a complete contraction of a tensor of even rank formed as a product of terms of the form $\partial^p f / \partial_{x_{i_1}} \cdots \partial_{x_{i_p}}$ will give a Euclidean invariant.

For example, up to second derivatives, we have the following Euclidean invariants, expressed as tensor contractions:
\begin{equation}
\label{eq:e2tensor}
\begin{aligned}
  I_0 &= f = f \\
  I_1 &= f_if_i = f_x^2 + f_y^2 \\
  I_2 &= f_{ii} = f_{xx} + f_{yy} \\
  I_3 &= f_{ij} f_{ij} = f_{xx}^2 + 2f_{xy}^2 + f_{yy}^2 \\
  I_4 &= f_i f_j f_{ij} = f_x^2 f_{xx} + 2f_x f_y f_{xy} + f_y^2 f_{yy}
\end{aligned}
\end{equation}

While we can construct any number of invariants this way, there is no
guarantee of functional independence between them. We will address this
issue in Section~\ref{sec:movingframes} when we construct invariants of
$E(2)$ and other groups using the method of moving frames. We can form a
signature from any three of these invariants \eqref{eq:e2tensor}, we choose for
simplicity our signature to be 
\begin{equation}
    \label{eq:e2signature}
    \mathcal{I}_{E(2)} = (I_0, I_1, I_2)
\end{equation}.  
A sample $E(2)$ transformation is depicted in Fig.~\ref{fig:E2} with its corresponding
signature.
\begin{figure}
\centering
\includegraphics[width=.45\textwidth]{Figs/f_transformed_E2.png}
\includegraphics[width=.45\textwidth]{Figs/E2_signature.png}
\caption{A sample $E(2)$ Transform and 3D signature.}
\label{fig:E2}
\end{figure}
We will defer discussion of the Special Euclidean group (which excludes
reflections) until Section~\ref{sec:SE2} where invariants naturally arise
from the method of moving frames.

\subsection{The similarity group}
\label{sec:Sim2}
The similarity group $\text{Sim}(2)$ consists transformations of the form
$\bar{\mathbf{x}} = \varphi(\mathbf{x}) = sU\mathbf{x} + \mathbf{t}$, where $U^TU = I$ and $s
\in \mathbb{R}\setminus\{0\}$. It is like the Euclidean group except it
also allows for isotropic scaling. In image applications, similarity
transformations arise from operations like zooming.

Under similarity transformations, the Euclidean invariants become
/emph{relative invariants}. Each derivative introduces a multiple of $s$ to
each invariant. For example, $\bar{f}_{\bar x}^2 + \bar{f}_{\bar y}^2 =
s^2(f_x^2 + f_y^2)$ and $\bar{f}_{\bar x \bar x}^2 + 2\bar{f}_{\bar x \bar
y}^2 + \bar{f}_{\bar y \bar y} = s^4 (f_{xx}^2 + 2f_{xy}^2 + f_{yy}^2$). We
refer to the power of $s$ as the weight of the transformation.

One way to produce similarity invariants would be to take ratios of powers
of the Euclidean invariants \eqref{eq:e2tensor} so that the multiples of
$s$ cancel. For example $\frac{f_x^2 + f_y^2}{f_{xx} + f_{yy}}$ is the
ratio of two weight-2 relative invariants and is hence a similarity
invariant. However this invariant is singular whenever $f_{xx} + f_{yy} =
0$ which is a codimension-0 phenomenon for images.

A better approach is to take three invariants of the same weight, we'll
take $I_1^2$, $I_2^2$, and $I_3$ from \eqref{eq:e2tensor}, all of which are
weight $4$, and project them onto the unit sphere. This can then be
multiplied by $I_0 = f$ to form a signature surface. So our signature is
defined by
\begin{equation}
    \label{eq:sim2signature}
\mathcal{I}_{\text{Sim(2)}}(f) = \frac{f}{\sqrt{I_1^4 + I_2^4 + I_3^2}}(I_1^2, I_2^2,
        I_3)
\end{equation}
This signature is only singular when $f_x = f_y = f_{xx} = f_{yy} = f_{xy}
= 0$. A sample similarity transformation and computed signature is shown in
Fig.~\ref{fig:Sim2}
\begin{figure}
\centering
\includegraphics[width=.45\textwidth]{Figs/f_transformed_Sim2.png}
\includegraphics[width=.45\textwidth]{Figs/Sim2_signature.png}
\caption{A sample $\text{Sim}(2)$ Transform and 3D signature.}
\label{fig:Sim2}
\end{figure}

@@

\subsection{The Special Euclidean Group SE(2)}
\label{sec:SE2}

@@

\begin{figure}
\centering
\includegraphics[width=.45\textwidth]{Figs/f_transformed_SE2.png}
\includegraphics[width=.45\textwidth]{Figs/SE2_signature.png}
\caption{A sample $SE(2)$ Transform and 3D signature.}
\label{fig:SE2}
\end{figure}
\subsection{The Euclidean Group E(2)}
An element of the Euclidean group $E(2)$ maps an element $(x, y)$ of
$\mathbb{R}^2$ to $(\bar{x}, \bar{y})$ as:
\begin{equation*}
  \begin{aligned}
    \bar{x} &=  x\varepsilon\cos\theta  - y\varepsilon\sin\theta + t_x \\
    \bar{y} &= x\sin\theta  + y\cos\theta + t_y.
  \end{aligned}
\end{equation*}
where $\varepsilon \in \{-1, 1\}$ and the remaining parameters are the same
as for $SE(2)$. If $\varepsilon = 1$ it is a rigid transformation, and if
$\varepsilon = -1$ it contains a reflection.
